---
title: "DATA 607 Project 1"
author: "Ilya Kats"
date: "February 24, 2017"
params:
  username: "root"
  password: ""
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project 1: Tournament Data

### Setup and Data Import
```{r libraries, message=FALSE}
# Load packages
library(stringr)
library(data.table)

# Read in data file skipping header block
tournament <- read.csv("C:\\Temp\\GitHub\\CUNY-DATA607\\tournamentinfo.txt", sep = ",", skip = 3)
```

### Data Parsing: Part 1

```{r data-parsing1}
# Remove all lines not containing pipes
tournament <- tournament[str_detect(tournament[,1], "\\|"), ]

# Combine first and second lines of each entry into a single line
tournament <- str_c(tournament[seq(1, length(tournament), 2)], tournament[seq(2, length(tournament), 2)], sep = "")

# Split into columns by pipes and at the same time split ratings
# Trim any whitespaces
tournament <- data.frame(str_split_fixed(tournament, "\\||/|->", n = Inf))
tournament <- data.frame(lapply(tournament, trimws))

# Drop last column
tournament <- tournament[, 1:ncol(tournament)-1]

# Change column names to meaningful values
colnames(tournament) <- c("playerId", "name", "points", 
                          "game1", "game2", "game3", "game4", "game5", "game6", "game7",
                          "state", "uscfId" , "ratingPre", "ratingPost", "nValue",
                          "color1", "color2", "color3", "color4", "color5", "color6", "color7")

# Normalize rating pre match - Remove 'R:' and provisional value
tournament$ratingPre <- as.numeric(str_replace(tournament$ratingPre, "R: *([:digit:]+)(P.*)?", "\\1"))

# Normalize rating post match - Remove provisional value
tournament$ratingPost <- as.numeric(str_replace(tournament$ratingPost, "([:digit:]+)(P.*)?", "\\1"))

# Convert ID to numeric
tournament$playerId <- as.numeric(as.character(tournament$playerId))
```

*Note: The table below lists only results of the first game.*

```{r, echo=FALSE}
knitr::kable(tournament[c(1:4,11:14, 16)], 
             col.names = c("ID", "Name", "Points", "Game 1", "State", "USCF ID", 
                           "Rating Pre", "Rating Post", "Color 1"),
             caption = "Table 1. Main data table.")
```

### Data Parsing: Part 2

```{r data-parsing2}
# Melt tournament data to work on individual games data
games <- melt(tournament, 
              id=c("playerId", "name"), 
              measure=c("game1", "game2", "game3", "game4", "game5", "game6", "game7"))

# Split results into result value and opponent id
games <- cbind(games[, 1:3],data.frame(str_split_fixed(games[, 4], " +", n = Inf)))

# Assign friendly names
colnames(games) <- c("playerId", "name", "gameNo", "result", "opponentId")

# Clean up game number column
games$gameNo <- str_replace(games$gameNo, "game", "")
```

```{r, echo=FALSE}
knitr::kable(head(games), 
             col.names = c("Player ID", "Name", "Game", "Result", "Opponent ID"),
             caption = "Table 2. Sample of individual games information.")
```

### Generate Opponent's Average Pre-Tournament Rating

```{r}
# Merge tournament and games information
# The merge drops any games without a W/L/D decision (opponent ID is blank)
games_expand <- data.table(merge(games, tournament, by.x = "opponentId", by.y = "playerId"))

# Calculate average pre-tournament rating
avg_pre_rating <- games_expand[, mean(ratingPre), by = playerId]

# Data cleanup 
colnames(avg_pre_rating) <- c("playerId", "oppAverage")
avg_pre_rating$oppAverage <- round(avg_pre_rating$oppAverage)

# Add average to the tournament data
tournament <- merge(tournament, avg_pre_rating, by = "playerId")
#tournament$playerId <- as.numeric(tournament$playerId)
#tournament <- tournament[order(tournament$points, tournament$playerId),]
#tournament
```

```{r, echo=FALSE}
knitr::kable(tournament[c(1, 2, 3, 11, 13, 23)], 
             col.names = c("ID", "Name", "Points", "State", "Pre-Tournament Rating", "Opp Avg Rating"),
             caption = "Table 3. Target values.")
```



Header block to be exactly 3 lines
Each entry to have 2 lines

W=Win
L=Loss
D=Draw
X=Win by forfeit (opponent failed to appear)
F=Loss by forfeit (player failed to appear)
U=Unplayed game
H=Half point bye
B=Full point bye