---
title: "DATA 607 Week 7 Assignment: HTML, XML, JSON"
author: "Ilya Kats"
date: "March 18, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<STYLE>
body {
    font-size: 18px;
}
table {
    border: 1px solid black;
    font-size: 14px;
}
th {
    background-color: rgb(112, 196, 105);
    color: white;
    font-weight: bold;
    padding: 20px 30px;
}
tr:nth-child(even) {
    background-color: rgb(220,220,220);
}
tr:nth-child(odd) {
    background-color: rgb(255, 255, 255);
}
</STYLE>

## Task

For this assignment I have created a listing of several of my favorite science fiction books in three formats - HTML table, XML and JSON. All files were created by hand in a text editor. All have just one author; however, I have added the awards information to satisfy the requirement of having multiple values in a field. The task is to import all files and convert them to data frames. Original file content and all R code is below.

## Required Libraries

```{r requirements, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
library(XML)       # For HTML and XML manipulation
library(RCurl)
library(tidyjson)  # For JSON manipulation
```

## HTML

### Original File

```{r comment="", echo=FALSE, warning=FALSE}
cat(readLines("https://raw.githubusercontent.com/ilyakats/CUNY-DATA607/master/books.html"), sep = "\n")
```

### Data Import and Manipulation

```{r}
# Import HTML file
htmlFile <- getURL("https://raw.githubusercontent.com/ilyakats/CUNY-DATA607/master/books.html")

# Parse HTML table
booksHTML <- readHTMLTable(htmlFile, header = TRUE)

# Convert to data frame and adjust column names
booksHTML <- as.data.frame(booksHTML)
colnames(booksHTML) <- c("title", "author", "year", "isbn", "award", "film")

# Adjust award column
booksHTML$award <- str_replace_all(booksHTML$award, "\t", "")
booksHTML$award <- str_replace_all(booksHTML$award, "\n", "; ")
```

### Converted Data Frame

```{r echo=FALSE}
knitr::kable(booksHTML)
```

The data was loaded and converted to a data frame with minimal manipulation (just minor clean-up). Of course, there is still a question about what to do with multiple award values for a given book. Film attributes can also be split accross multiple columns.

## XML

### Original File

```{r comment="", echo=FALSE, warning=FALSE}
cat(readLines("https://raw.githubusercontent.com/ilyakats/CUNY-DATA607/master/books.xml"), sep = "\n")
```

### Data Import and Manipulation

```{r}
# Import XML file
xmlFile <- getURL("https://raw.githubusercontent.com/ilyakats/CUNY-DATA607/master/books.xml")

# Parse XML and get root
booksXML <- xmlParse(xmlFile)
root <- xmlRoot(booksXML)

# Convert main fields to data frame
booksXML <- xmlToDataFrame(root)

# Loop through XML nodes to get film attributes
for (i in 1:xmlSize(root)) {
  if (!is.null(root[[i]][["film"]])) { 
    booksXML[i, "film"] <- paste(xmlAttrs(root[[i]][["film"]]), collapse = " - ")
  }
}

# Loop thorugh XML nodes to collect award data
for (i in 1:xmlSize(root)) {
  if (!is.null(root[[i]][["awardlist"]])) { 
    awards <- xmlChildren(root[[i]][["awardlist"]])
    awardlist <- ""
    for (j in 1:xmlSize(awards)) {
      jaward <- paste(xmlValue(awards[[j]][["name"]]), xmlValue(awards[[j]][["year"]]), sep = " - ")
      if (j==1) {sep <- ""} else {sep <- "; "}
      awardlist <- paste(awardlist, jaward, sep = sep)
    }
    booksXML[i, "awardlist"] <- awardlist
  }
}


```

### Converted Data Frame

```{r echo=FALSE}
knitr::kable(booksXML)
```

XML format required more work. Main fields were easily converted, but it took a couple of loops to traverse the XML tree structure to gather film attributes and award information. For this exercise, multiple awards were merged into a single entry similar to HTML output.

## JSON

### Original File

```{r comment="", echo=FALSE, warning=FALSE}
cat(readLines("https://raw.githubusercontent.com/ilyakats/CUNY-DATA607/master/books.json"), sep = "\n")
```

### Data Import and Manipulation

```{r}
# Import JSON file
jsonFile <- getURL("https://raw.githubusercontent.com/ilyakats/CUNY-DATA607/master/books.json")

# Parse book data
booksJSON <- jsonFile %>% 
  enter_object("scifi_books") %>% 
  gather_array("book.id") %>% 
  spread_values(title = jstring("title"), 
                author = jstring("author"),
                year = jnumber("year"), 
                isbn = jstring("isbn"))

# Parse award data
awards <- jsonFile %>% 
  enter_object("scifi_books") %>% 
  gather_array("book.id") %>%                    # Get id for all books to join later
  enter_object("award") %>%                      # Get a listing of all awards
  gather_array() %>% 
  spread_values(award.name = jstring("name"),
                award.year = jnumber("year"))

# Parse film data
films <- jsonFile %>% 
  enter_object("scifi_books") %>% 
  gather_array("book.id") %>%                    # Get id for all books to join later
  enter_object("film") %>%                       # Get a listing of all films
  gather_array() %>% 
  spread_values(film.title = jstring("name"),
                film.director = jstring("director"),
                film.year = jnumber("year"))

# Merge book, award and film data
booksJSONflat <- booksJSON %>% 
  left_join(awards, by = "book.id") %>% 
  left_join(films, by = "book.id") %>% 
  select(book.id, title, author, year, isbn, award.name, award.year, 
         film.title, film.director, film.year)
```

### Converted Data Frame

```{r echo=FALSE}
knitr::kable(booksJSONflat)
```

With some research `tidyjson` package turned out to be very powerful in manipulating JSON data. For this exercise I went with a long structure for the data frame based on award data. 

## Conclusion

It took quite a bit of time to figure out how to traverse different data structures. XML and JSON allow for the most flexibility in manipulating stored data. HTML is simplier to handle, but much less flexible.