---
title: "DATA 607 Week 2 Assignment"
author: "Ilya Kats"
date: "February 11, 2017"
params:
  username: "root"
  password: ""
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 2 task 

Choose six recent popular movies. Ask at least five people that you know (friends, family, classmates, imaginary friends) to rate each of these movie that they have seen on a scale of 1 to 5. Take the results (observations) and store them in a SQL database. Load the information into an R dataframe. Your deliverables should include your SQL scripts and your R Markdown code, posted to GitHub.

## Setup

```{r libraries, message=FALSE}
# Load packages
library(RMySQL)
library(reshape2)
library(plyr)
```

## Data initialization

Since this assignment is centered around using R with SQL, I have decided to enter initial data manually. Every year I attemp to watch all movies nominated as Best Picture for the Academy Awards. This data set includes my ratings and ratings of several review aggregators for 9 movies nominated this year. 

```{r data_initialization}

movies <- c("Arrival", "Fences", "Hacksaw Ridge", "Hell or High Water", "Hidden Figures",
            "La La Land", "Lion", "Manchester by the Sea", "Moonlight")
ilya <- c(3, NA, 3, NA, 4, 3, 3, 5, 4)
imdb <- c(8.1, 7.5, 8.3, 7.7, 7.9, 8.5, 8, 8.2, 8.1)
rotten_tomatoes <- c(94, 95, 86, 98, 92, 93, 86, 96, 98)
google_users <- c(86, 83, 94, 91, 92, 89, 84, 74, 88)
metacritic <- c(81, 79, 71, 88, 74, 93, 69, 96, 99)

ratings <- data.frame(movies, ilya, imdb, rotten_tomatoes, google_users, metacritic)

# Normalize all ratings to a 5 point scale
ratings$imdb <- round(ratings$imdb / 2, 0)
ratings$rotten_tomatoes <- round(ratings$rotten_tomatoes / 20, 0)
ratings$google_users <- round(ratings$google_users / 20, 0)
ratings$metacritic <- round(ratings$metacritic / 20, 0)
knitr::kable(ratings, col.names = c("Movies", "My Ratings", "IMDB", "Rotten Tomatoes", "Google Users", "Metacritic"))
```

Just for practicing I have decided to normalize the main table into three smaller tables containing information about movies, critics and ratings. 

```{r data_transpose}
# Transpose the table
movie_ratings <- melt(ratings, id=c("movies"))
colnames(movie_ratings) <- c("movie", "critic", "rating")

# Extract unique movies and assign ID to each
movies <- data.frame(1:nlevels(movie_ratings$movie), levels(movie_ratings$movie))
colnames(movies) <- c("id", "movie")

# Extract unique critics and assign ID to each
critics <- data.frame(1:nlevels(movie_ratings$critic), levels(movie_ratings$critic))
colnames(critics) <- c("id", "critic")

# Add corresponding movie and critic IDs to ratings
movie_ratings <- join(movie_ratings, critics, by = "critic")
movie_ratings <- join(movie_ratings, movies, by = "movie")
# Drop unneeded columns
movie_ratings <- movie_ratings[ -c(1:2)]
# Add ID to ratings and adjust column names
movie_ratings <- data.frame(1:nrow(movie_ratings), movie_ratings)
colnames(movie_ratings) <- c("id", "rating", "critic_id", "movie_id")

# Adjust critic names
critics$critic <- as.character(critics$critic)
critics$critic[critics$critic == 'ilya'] <- 'Ilya Kats'
critics$critic[critics$critic == 'imdb'] <- 'IMDb'
critics$critic[critics$critic == 'rotten_tomatoes'] <- 'Rotten Tomatoes'
critics$critic[critics$critic == 'google_users'] <- 'Google Users'
critics$critic[critics$critic == 'metacritic'] <- 'Metacritic'
critics$critic <- as.factor(critics$critic)
```

```{r echo=FALSE}
knitr::kable(movies, col.names = c("ID", "Movie"), caption = "Table 1. Movies")
knitr::kable(critics, col.names = c("ID", "Critic"), caption = "Table 2. Critics")
knitr::kable(movie_ratings[1:12, ], col.names = c("ID", "Rating", "Critic ID", "Movie ID"), caption = "Table 3. Ratings")
```

## Database connection

```{r db_connection, eval=FALSE}
# Prompt for server username and password and establish connection
# getPass package is required

username <- readline("Please enter MySQL server username: ")
password <- getPass::getPass("Please enter MySQL server password: ")

con <- dbConnect(MySQL(), "flights", username=username, password=password)
```

```{r db_connection_markdown, echo=FALSE}
# Establish connection for rendering of R Markdown document by passing username and password through parameters
# Do not execute outside of the R Markdown document rendering

con <- dbConnect(MySQL(), "flights", username=params$username, password=params$password)
```

## Database creation, setup and data import

```{r data_import, results='hide'}
dbSendQuery(con, "CREATE DATABASE IF NOT EXISTS MovieRatings;")
dbSendQuery(con, "USE MovieRatings;")

# MOVIES
dbSendQuery(con, "DROP TABLE IF EXISTS Movies;")
dbSendQuery(con, "CREATE TABLE Movies 
            (
            ID int PRIMARY KEY, 
            Title varchar(255)
            );")
values <- paste("(",movies$id,",'", movies$movie,"')", sep="", collapse=",")
sqlst <- paste("INSERT INTO Movies VALUES",values)
dbSendQuery(con, sqlst)

# CRITICS
dbSendQuery(con, "DROP TABLE IF EXISTS Critics;")
dbSendQuery(con, "CREATE TABLE Critics 
            (
            ID int PRIMARY KEY, 
            Name varchar(255)
            );")
values <- paste("(",critics$id,",'", critics$critic,"')", sep="", collapse=",")
sqlst <- paste("INSERT INTO Critics VALUES",values)
dbSendQuery(con, sqlst)

# RATINGS
dbSendQuery(con, "DROP TABLE IF EXISTS Ratings;")
dbSendQuery(con, "CREATE TABLE Ratings 
            (
            ID int PRIMARY KEY, 
            CriticID int,
            MovieID int,
            Rating int
            );")
values <- paste("(",movie_ratings$id,",", movie_ratings$critic_id,",", 
                movie_ratings$movie_id,",", 
                ifelse(!is.na(movie_ratings$rating),movie_ratings$rating,"NULL"),
                ")", sep="", collapse=",")
sqlst <- paste("INSERT INTO Ratings VALUES",values)
dbSendQuery(con, sqlst)

# Create a view to easily retrieve all relevant ratings information
dbSendQuery(con, "DROP VIEW IF EXISTS ExpandedRatings;")
sqlst <- "CREATE VIEW ExpandedRatings AS
          SELECT c.Name AS 'Critic Name', m.Title AS 'Movie Title', r.Rating FROM Ratings r
          INNER JOIN Critics c ON r.CriticID = c.ID
          INNER JOIN Movies m ON r.MovieID = m.ID;"
dbSendQuery(con, sqlst)
```

## Data access

```{r data_access}
allratings <- dbReadTable(con, "ExpandedRatings")
knitr::kable(allratings)
```

Other options for reading data from SQL database into R dataframe are `dbGetQuery` to implicitly fetch all results of a SELECT statement and `dbSendQuery` and `fetch` combination to access results, but only fetch them when explicitly specified.

## Cleanup

```{r cleanup, results='hide'}
dbDisconnect(con)
```

## Conclusion

The code above connects to a MySQL server and creates a simple relational database completely by passing SQL statements through R code. Alternatively, data could have been saved in an external CSV file and loaded into a SQL database by executing the LOAD DATA SQL statement. 

```sql
LOAD DATA INFILE 'c:/Data/ratings.csv' 
INTO TABLE ratings 
FIELDS TERMINATED BY ',' 
ENCLOSED BY '"'
LINES TERMINATED BY '\n'
IGNORE 1 ROWS;
```