---
title: "DATA 607 Week 1 Assignment"
author: "Ilya Kats"
date: "February 5, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 1 task 

**Your task** (***should you choose to except it** -- sorry, had the Mission Impossible theme playing as soon as I started reading*) is to study the dataset and the associated description of the data (i.e. "data dictionary"). You may need
to look around a bit, but it's there! You should take the data, and create a data frame with a subset of the columns
in the dataset. You should include the column that indicates edible or poisonous and three or four other columns.
You should also add meaningful column names and replace the abbreviations used in the data-for example, in the
appropriate column, "e" might become "edible." Your deliverable is the R code to perform these transformation
tasks.

## Setup

The data set includes descriptions of hypothetical samples corresponding to 23 species of gilled mushrooms in the Agaricus and Lepiota Family. Data is courtesy of [UCI](https://archive.ics.uci.edu/ml/datasets/Mushroom/), Jeff Schlimmer and The Audubon Society. 

```{r message=FALSE}
# Load packages
library(RCurl)

# Load data file
shrooms <- read.csv(text=getURL("https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.data"), header = FALSE, sep = ",")
```

```{r}
# Quick look at the data
head(shrooms)
summary(shrooms)
```

## Data dictionary

The first column describes class (edible=e, poisonous=p) and the data set includes 22 variables:

1. `cap-shape`: bell=b, conical=c, convex=x, flat=f, knobbed=k, sunken=s
2. `cap-surface`: fibrous=f, grooves=g, scaly=y, smooth=s
3. `cap-color`: brown=n, buff=b, cinnamon=c, gray=g, green=r, pink=p, purple=u, red=e, white=w, yellow=y
4. `bruises?`: bruises=t, no=f
5. `odor`: almond=a, anise=l, creosote=c, fishy=y, foul=f, musty=m, none=n, pungent=p, spicy=s
6. `gill-attachment`: attached=a, descending=d, free=f, notched=n
7. `gill-spacing`: close=c, crowded=w, distant=d
8. `gill-size`: broad=b, narrow=n
9. `gill-color`: black=k, brown=n, buff=b, chocolate=h, gray=g, green=r, orange=o, pink=p, purple=u, red=e, white=w, yellow=y
10. `stalk-shape`: enlarging=e, tapering=t
11. `stalk-root`: bulbous=b, club=c, cup=u, equal=e, rhizomorphs=z, rooted=r, missing=?
12. `stalk-surface-above-ring`: fibrous=f, scaly=y, silky=k, smooth=s
13. `stalk-surface-below-ring`: fibrous=f, scaly=y, silky=k, smooth=s
14. `stalk-color-above-ring`: brown=n, buff=b, cinnamon=c, gray=g, orange=o, pink=p, red=e, white=w, yellow=y
15. `stalk-color-below-ring`: brown=n, buff=b, cinnamon=c, gray=g, orange=o, pink=p, red=e, white=w, yellow=y
16. `veil-type`: partial=p, universal=u
17. `veil-color`: brown=n, orange=o, white=w, yellow=y
18. `ring-number`: none=n, one=o, two=t
19. `ring-type`: cobwebby=c, evanescent=e, flaring=f, large=l, none=n, pendant=p, sheathing=s, zone=z
20. `spore-print-color`: black=k, brown=n, buff=b, chocolate=h, green=r, orange=o, purple=u,white=w, yellow=y
21. `population`: abundant=a, clustered=c, numerous=n, scattered=s, several=v, solitary=y
22. `habitat`: grasses=g, leaves=l, meadows=m, paths=p, urban=u, waste=w, woods=d

```{r}
# Rename columns 
names(shrooms) <- c('class','capshape','capsurface','capcolor','bruises','odor','gillattachment','gillspacing','gillsize','gillcolor','stalkshape','stalkroot','stalksurfaceabovering','stalksurfacebelowring','stalkcolorabovering','stalkcolorbelowring','veiltype','veilcolor','ringnumber','ringtype','sporeprintcolor','population','habitat')
```

## Selecting a subset

Select `class`, `odor`, `ringnumber`, `ringtype` and `population` for all mushrooms that grows in the woods to further transform and review.

```{r}
sub.shrooms <- subset(shrooms, habitat == 'd', select = c(class, odor, ringnumber, ringtype, population))
summary(sub.shrooms)
```

## Transforming the subset

Update variables with more meaningful values.

```{r}
# Update class
sub.shrooms$class <- as.character(sub.shrooms$class)
sub.shrooms$class[sub.shrooms$class == 'e'] <- 'Edible'
sub.shrooms$class[sub.shrooms$class == 'p'] <- 'Poisonous'
sub.shrooms$class <- as.factor(sub.shrooms$class)

# Update odor
sub.shrooms$odor <- as.character(sub.shrooms$odor)
sub.shrooms$odor[sub.shrooms$odor == 'a'] <- 'Almond'
sub.shrooms$odor[sub.shrooms$odor == 'l'] <- 'Anise'
sub.shrooms$odor[sub.shrooms$odor == 'c'] <- 'Creosote'
sub.shrooms$odor[sub.shrooms$odor == 'y'] <- 'Fishy'
sub.shrooms$odor[sub.shrooms$odor == 'f'] <- 'Foul'
sub.shrooms$odor[sub.shrooms$odor == 'm'] <- 'Musty'
sub.shrooms$odor[sub.shrooms$odor == 'n'] <- 'None'
sub.shrooms$odor[sub.shrooms$odor == 'p'] <- 'Pungent'
sub.shrooms$odor[sub.shrooms$odor == 's'] <- 'Spicy'
sub.shrooms$odor <- as.factor(sub.shrooms$odor)

# Update ring type; Replace with NA if no rings exist
sub.shrooms$ringtype <- as.character(sub.shrooms$ringtype)
sub.shrooms$ringtype[sub.shrooms$ringtype == 'p'] <- 'Pendant'
sub.shrooms$ringtype[sub.shrooms$ringtype == 'c'] <- 'Cobwebby'
sub.shrooms$ringtype[sub.shrooms$ringtype == 'e'] <- 'Evanescent'
sub.shrooms$ringtype[sub.shrooms$ringtype == 'f'] <- 'Flaring'
sub.shrooms$ringtype[sub.shrooms$ringtype == 'l'] <- 'Large'
sub.shrooms$ringtype[sub.shrooms$ringtype == 'n'] <- NA
sub.shrooms$ringtype[sub.shrooms$ringtype == 's'] <- 'Sheathing'
sub.shrooms$ringtype[sub.shrooms$ringtype == 'z'] <- 'Zone'
sub.shrooms$ringtype <- as.factor(sub.shrooms$ringtype)

# Update population
sub.shrooms$population <- as.character(sub.shrooms$population)
sub.shrooms$population[sub.shrooms$population == 'a'] <- 'Abundant'
sub.shrooms$population[sub.shrooms$population == 'c'] <- 'Clustered'
sub.shrooms$population[sub.shrooms$population == 'n'] <- 'Numerous'
sub.shrooms$population[sub.shrooms$population == 's'] <- 'Scattered'
sub.shrooms$population[sub.shrooms$population == 'v'] <- 'Several'
sub.shrooms$population[sub.shrooms$population == 'y'] <- 'Solitary'
sub.shrooms$population <- as.factor(sub.shrooms$population)

# Update ring number and convert to numeric
sub.shrooms$ringnumber <- as.character(sub.shrooms$ringnumber)
sub.shrooms$ringnumber[sub.shrooms$ringnumber == 'n'] <- 0
sub.shrooms$ringnumber[sub.shrooms$ringnumber == 'o'] <- 1
sub.shrooms$ringnumber[sub.shrooms$ringnumber == 't'] <- 2
sub.shrooms$ringnumber <- as.numeric(sub.shrooms$ringnumber)

# Display 25 random rows to check the transformations
set.seed(125)
sub.shrooms[sample(1:nrow(sub.shrooms), 25), ]

summary(sub.shrooms)
```

Adding a few rudimentary graphs just for a bit of practice.

```{r}
plot(sub.shrooms$class ~ sub.shrooms$ringtype, xlab = "Ring Type", ylab = "Class", main = "Class by Ring Type (Woods Only)")
plot(sub.shrooms$odor, xlab = "Odor", ylab = "Frequency", col = 1:length(sub.shrooms$odor))
```